### 2014-10-02 Paper_v1.tex
New network fig, mm equations and intro

\section*{Introduction}
Developing mathematical models of biochemical networks is a significant facet of systems biology.
Modeling approaches differ in their degree of detail, and the choice of approach is often determined by prior system knowledge \citep{2012_kholodenko_kolch_SciSig}. 
The ability of ordinary differential equations (ODEs) to capture dynamics has made it one of the most common tools for studying signaling systems. 
However, ODEs come with difficult (and sometimes impossible) parameter identification problems. 
For example, Gadkar et al., showed that even with near-perfect information and high frequency sampling, 
it was often impossible to identify (to within a specified uncertainty) all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
This result highlights the perhaps under-appreciated role that experimental design could play in generating the best training and validation data sets for model identification \citep{Apgar:2010fk}.
ODE models also require significant mechanistic knowledge of the underlying biology. 
However, Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers showed that model behavior and predictive ability are typically 
sensitive to only a few parameter combinations, a characteristic seemingly universal to multi-parameter signaling models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
Alternatively, data-driven approaches \citep{Cirit:2012kx} or logical model formulations \citep{Morris:2011ys} 
are emerging paradigms that constrain model structure (and hence the identifiability challenge) by the availability of training and validation data. 
These are interesting techniques because the data naturally informs model structure, which can be expanded as more data becomes available. 
Thus, while there are many possible strategies to simulate biochemical networks, ODE models remain a common choice despite
significant identification challenges.

The human coagulation system is an archetype proteolytic cascade regulated by both positive and negative feedback loops.
The biology of coagulation is well studied \cite{ROBERTS1998,MANN1999,BUTENAS2002,FURIE2004}. 
Coagulation is mediated by a family proteases, called factors and a key group of blood cells, called platelets. 
The activation of coagulation factors and platelets, which are normally inactive in the circulation, requires a trigger event.
Trigger events such as vessel injury expose procoagulant materials like collagen, fibronectin, von Willebrand factor (vWF) or tissue factor (TF).
These materials drive platelet localization and activation, coagulation factor activation and ultimately clot formation \cite{GIESEN1999}.  
Once coagulation has been activated, two converging pathways activate the master protease thrombin. 
The extrinsic cascade is generally believed to be the main mechanism of thrombinogenesis in the blood \cite{MANN1990,ROBERTS1998,MANN1999}.
Thrombin generation consists of three phases, initiation, propagation and termination \cite{GOLDHABER2006, BRUMMEL2002}.  
Initially thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa) \cite{BUTENAS2002a}. 
Picomolar amounts of thrombin can then activate the cofactors factors V and VIII (fV and fVIII) and platelets, 
resulting in the formation of the \emph{tenase} and \emph{prothrombinase} complexes on the surface of activated platelets.
Tenese amplifies the early coagulation signal by activating FXa, while prothrombinase, a complex of FVa-FXa, 
directly converts prothrombin to thrombin. There are several control points in the cascade that promote 
or inhibit thrombin formation. Tissue Factor Pathway Inhibitor (TFPI) downregulates FXa formation catalyzed by TF/FVIIa, while antithrombin III (ATIII) neutralizes 
several of the serine proteases generated during coagulation.  
Thrombin itself also inadvertently plays a role in its own inhibition; thrombin, through interaction with thrombomodulin and endothelial cell protein C receptor (EPCR),
converts protein C to activated protein C (APC) which attenuates the coagulation response by the proteolytic cleavage of fV/FVa and fVIII/FVIIIa. 
Thus, termination occurs after prothrombin is consumed or activated thrombin is neutralized by inhibitors such as activated protein C (APC) or antithrombin III (ATIII).

@@

\subsection*{Identification of model parameters using particle swarm optimization.}
A critical challenge for any dynamic model is the estimation of kinetic parameters. 
We estimated kinetic and control parameters simultaneously from \emph{in vitro} coagulation data sets with and without the protein C pathway. 
The residual between model simulations and experimental measurements was minimized using particle swarm optimization (PSO). 
A population of particles (N = 20) was initialized with randomized kinetic and control parameters and allowed to search for parameter vectors that minimized
the residual. However, not all parameters were varied simultaneously. 
We partitioned the parameter estimation problem into two subproblems based upon the biological organization of the training data; 
(i) estimation of parameters associated with thrombin formation in the absence of the protein C pathway and (ii) estimation of parameters associated with the protein C pathway.
Only those parameters associated with each subproblem were varied during the optimization procedure for that subproblem e.g., thrombin parameters were \textit{not} varied during the protein C subproblem.
The PSO procedure was run for 20 generations for each subproblem, where each generation was 1200 iterations. The best particle from each generation was used to generate the particle population
for the next generation. We rotated the subproblems, starting with subproblem 1 in the first generation.

@@

\subsection*{Formulation and solution of the model equations.}
We used ordinary differential equations (ODEs) to model the time evolution of proteins ($x_{i}$) in our reduced order coagulation model:
\begin{equation}
	\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
\end{equation}
where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of protein species in the model. 
The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$. 
Typically, reaction $j$ is a non-linear function of biochemical species abundance, as well as unknown kinetic parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$). 
The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. 
If $\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. 
Conversely, if $\sigma_{ij}>0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates metabolite $i$ is not connected with reaction $j$. 
Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant. 
The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$.

@@

The control term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
For each enzyme, we used a rule-based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
If an enzyme was activated by $m$ metabolites, we modeled this activation as:
\begin{equation}
	v_{j} = \max\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right)
\end{equation}where $0\leq f_{ij}\left(\mathcal{Z}\right)\leq 1$ was a regulatory transfer function that calculated the influence of metabolite $i$ on the activity of enzyme $j$. 
Conversely, if enzyme activity was inhibited by a $m$ metabolites, we modeling this inhibition as:
\begin{equation}
	v_{j} = 1 - \max\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right)
\end{equation}Lastly, if an enzyme had both $m$ activating and $n$ inhibitory factors, we modeled the regulatory term as:
\begin{equation}
	v_{j} = \min\left(u_{j},d_{j}\right)
\end{equation}where:
\begin{eqnarray}
	u_{j} &=& \max_{j^{+}}\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right) \\
	d_{j} &=& 1 - \max_{j^{-}}\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{nj}\left(\mathcal{Z}\right)\right)
\end{eqnarray}
The quantities $j^{+}$ and $j^{-}$ denoted the sets of activating and inhibitory factors for enzyme $j$. 
If a process has no modifying factors, we set $v_{j} = 1$.
There are many possible functional forms for $0\leq f_{ij}\left(\mathcal{Z}\right)\leq 1$. 
However, in this study, each individual transfer function took the form:
\begin{equation}\label{eqn:control-factor}
	f_{i}\left(\mathbf{x}\right) = \frac{\kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}
\end{equation}where $\mathcal{Z}_{j}$ denotes the abundance of the $j$ factor (e.g., metabolite abundance), and $\kappa_{ij}$ and $\eta$ are control parameters. 
The $\kappa_{ij}$ parameter was species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
Applying the general framework to the reduced coagulation network resulted in five ordinary differential equations:
\begin{eqnarray}
	\frac{dx_{1}}{dt} & =& -\left(r_{init}v_{init} + r_{amp}v_{amp}\right)\\
	\frac{dx_{2}}{dt} &=& r_{amp}v_{amp} + r_{init}v_{init} - \left(r_{inh,ATIII}v_{inh,ATIII} + r_{inh,APC}v_{inh,APC}\right)\\
	\frac{dx_{3}}{dt} &=& -r_{apc}v_{apc} \\
	\frac{dx_{4}}{dt} &=& r_{apc}v_{apc} \\
	\frac{dx_{5}}{dt} & = & -r_{inh,ATIII}v_{inh,ATIII}
\end{eqnarray}where $\mathbf{x} = \left(fII,FIIa,PC,APC,ATIII\right)^{T}$.
The model equations were encoded using the Python programming language and solved using the ODEINT routine of the SciPy module \citep{SCIPY}.

\subsection*{Estimation of model parameters from experimental data.}
Model parameters were estimated by minimizing the difference between simulations and experimental thrombin measurements (squared residual):
\begin{equation}\label{eqn:objective-function}
	\min_{\mathbf{k}} \sum_{\tau=1}^{\mathcal{T}}\sum_{j=1}^{\mathcal{S}}\left(\frac{\hat{x}_{j}\left(\tau\right) - x_{j}\left(\tau,\mathbf{k}\right)}{\omega_{j}\left(\tau\right)}\right)^{2}
\end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated 
value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is respect to
time, while the inner summation is with respect to state. We minimized the model residual using Particle swarm optimization (PSO) \citep{PSO}.
PSO uses a \textit{swarming} metaheuristic to explore parameter spaces. 
A strength of PSO is its ability to find the global minimum, even in the presence of potentially many local minima, by communicating the local
error landscape experienced by each particle collectively to the swarm. Thus, PSO acts both as a local and a global search algorithm. 
For each iteration, particles in the swarm compute their local error by evaluating the model equations using their specific parameter vector realization.
From each of these local points, a globally best error is identified. Both the local and global error 
are then used to update the parameter estimates of each particle using the rules:
\begin{eqnarray}
	\mathbf{\Delta}_{i} &=&\theta_{1}\mathbf{\Delta}_{i} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{k}_{i}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G} - \mathbf{k}_{i}\right) \\
	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes local best solution found by particle $i$, and
$\mathcal{G}$ denotes the best solution found over the entire population of particles. The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$. The quality of parameter
estimates was measured using two criteria, goodness of fit (model residual) and angle between the estimated parameter vector $\mathbf{k}_{j}$ and the best parameter set $\mathbf{k}^{*}$:
\begin{equation}
	\alpha_{j} = cos^{-1}\left(\frac{\mathbf{k}_{j}\cdot{\mathbf{k^{*}}}}{\norm{\mathbf{k}_{j}}\norm{\mathbf{k^{*}}}}\right)
\end{equation}If the candidate parameter set $\mathbf{k}_{j}$ were perfect, the residual between the model and synthetic data 
and the angle between $\mathbf{k}_{j}$ and the true parameter set $\mathbf{k}^{*}$ would be equal to zero.
The particle swarm optimization routine was implemented in the Python programming language. 
All plots were made using the Matplotlib module of Python \citep{Matplotlib}.

@@

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-1-Network.pdf}
\caption{Schematic of the connectivity of the reduced order coagulation model. A trigger compound e.g., TF/FVIIa initiates thrombin production (FIIa) from prothrombin (fII).
Once activated, thrombin catalyzes its own activation (amplification step), as well as its own inhibition via the conversion of protein C to activated protein C (APC).
APC and tissue factor pathway inhibitor (TFPI) inhibit initiation, while antithrombin III (ATIII) directly inhibits thrombin. 
All inhibition steps, and trigger-induced initiation were modeled using a rule-based approach. 
Likewise, the dependence of amplification on other coagulation factors was also modeled using a rule-based approach. }\label{fig-network}
\end{figure}

>>>>

%
%Thus, while there are many possible strategies to simulate biochemical networks, ODE models remain a common choice despite
%significant identification challenges.
%This result highlights the perhaps under-appreciated role that experimental design could play in generating the best training and validation data sets for model identification \citep{Apgar:2010fk}.

\section*{Introduction}
Developing mathematical models of biochemical networks is a significant facet of systems biology.
Modeling approaches differ in their degree of detail. 
The choice of approach is often determined by prior system knowledge, and model requirements \citep{2012_kholodenko_kolch_SciSig}. 
Ordinary differential equation (ODE) models are a common tool for modeling biochemical systems because of their ability to capture dynamics. 
However, ODEs often come with difficult (and sometimes impossible) parameter identification problems. 
For example, Gadkar et al., showed that even with near-perfect information and high frequency sampling, 
it was often impossible to identify (to within a specified uncertainty) all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
ODE models can also require significant mechanistic knowledge of the underlying biology. 
However, Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers showed that model behavior and predictive ability is typically 
sensitive to only a few parameter combinations, a characteristic seemingly universal to multi-parameter signaling models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
Thus, accurate model predictions may be possible even if parameters or mechanistic structure is not well understood.
Alternatively, data-driven approaches \citep{Cirit:2012kx} or logical model formulations \citep{Morris:2011ys} 
are emerging paradigms that constrain model structure (and hence the identifiability challenge) by the availability of training and validation data. 
These are interesting techniques because the data naturally informs model structure, which can be expanded as more data becomes available. 
Thus, while not mechanistic, these models can simulate surprisingly rich system responses in the absence of mechanism. 
A potential third approach could be a combination of ODEs and logical models, where mechanistic information could be used when available, 
and missing or incomplete knowledge encoded using causal local rules.

In this study, we explored a hybrid modeling approach involving both ODEs and logical rules on a well studied biochemical network, 
the human coagulation system. The human coagulation system is an archetype proteolytic cascade involving both positive 
and negative feedback \cite{Butenas:2002aa,Schenone:2004aa,Adams:2009aa}. 
Coagulation is mediated by a family proteases in the circulation, called factors and a key group of blood cells, called platelets. 
The activation of coagulation factors and platelets, which are normally inactive in the circulation, requires a trigger event.
Trigger events such as vessel injury expose procoagulant materials like collagen, fibronectin, von Willebrand factor (vWF) or tissue factor (TF).
These materials drive platelet localization and activation, coagulation factor activation and ultimately clot formation \cite{GIESEN1999}.  
Once coagulation has been activated, two converging pathways activate the master protease thrombin (FIIa) from prothrombin (fII). 
The extrinsic cascade is generally believed to be the main mechanism of thrombinogenesis in the blood \cite{MANN1990,ROBERTS1998,MANN1999}.
Thrombin generation consists of three phases, initiation, amplification and termination \cite{GOLDHABER2006, BRUMMEL2002}.  
Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa) \cite{BUTENAS2002a}. 
Picomolar amounts of thrombin then activate the cofactors factors V and VIII (fV and fVIII) and platelets, 
resulting in the formation of the tenase and prothrombinase complexes on activated platelets.
Tenese amplifies the early coagulation signal by activating FXa, while prothrombinase, a complex of FVa-FXa, 
directly converts prothrombin to thrombin. There are several control points in the cascade that inhibit thrombin formation. 
Tissue Factor Pathway Inhibitor (TFPI) downregulates FXa formation catalyzed by TF/FVIIa, while antithrombin III (ATIII) \
neutralizes several of the serine proteases generated during coagulation.  
Thrombin itself also inadvertently plays a role in its own inhibition; thrombin, through interaction with thrombomodulin, protein C and endothelial cell protein C receptor (EPCR),
converts protein C to activated protein C (APC) which attenuates the coagulation response by proteolytic cleavage of fV/FVa and fVIII/FVIIIa. 
Termination occurs after prothrombin is consumed or activated thrombin formation is neutralized by inhibitors such as activated protein C (APC) or antithrombin III (ATIII).

@@

\subsection*{Identification of model parameters using particle swarm optimization.}
A critical challenge for any dynamic model is the estimation of kinetic parameters. 
We estimated kinetic and control parameters simultaneously from eight \emph{in vitro} time-series coagulation data sets with and without the protein C pathway. 
The residual between model simulations and experimental measurements was minimized using particle swarm optimization (PSO). 
A population of particles (N = 20) was initialized with randomized kinetic and control parameters and allowed to search for parameter vectors that minimized
the residual. However, not all parameters were varied simultaneously. 
We partitioned the parameter estimation problem into two subproblems based upon the biological organization of the training data; 
(i) estimation of parameters associated with thrombin formation in the absence of the protein C pathway and (ii) estimation of parameters associated with the protein C pathway.
Only those parameters associated with each subproblem were varied during the optimization procedure for that subproblem e.g., thrombin parameters were \textit{not} varied during the protein C subproblem.
The PSO procedure was run for 20 generations for each subproblem, where each generation was 1200 iterations. The best particle from each generation was used to generate the particle population
for the next generation. We rotated the subproblems, starting with subproblem 1 in the first generation.

@@

\subsection*{Formulation and solution of the model equations.}
We used ordinary differential equations (ODEs) to model the time evolution of proteins ($x_{i}$) in our reduced order coagulation model:
\begin{equation}
	\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
\end{equation}
where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of protein species in the model. 
The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$. 
Typically, reaction $j$ is a non-linear function of biochemical species abundance, as well as unknown kinetic parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$). 
The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. 
If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$. 
Conversely, if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates species $i$ is not connected with reaction $j$. 
The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$, which were specified by the experimental setup.

@@

The control term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
For each enzyme, we used a rule-based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
If an enzyme was activated by $m$ metabolites, we modeled this activation as:
\begin{equation}
	v_{j} = \max\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right)
\end{equation}where $0\leq f_{ij}\left(\mathcal{Z}\right)\leq 1$ was a regulatory transfer function that calculated the influence of metabolite $i$ on the activity of enzyme $j$. 
Conversely, if enzyme activity was inhibited by a $m$ metabolites, we modeling this inhibition as:
\begin{equation}
	v_{j} = 1 - \max\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right)
\end{equation}Lastly, if an enzyme had both $m$ activating and $n$ inhibitory factors, we modeled the regulatory term as:
\begin{equation}
	v_{j} = \min\left(u_{j},d_{j}\right)
\end{equation}where:
\begin{eqnarray}
	u_{j} &=& \max_{j^{+}}\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right) \\
	d_{j} &=& 1 - \max_{j^{-}}\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{nj}\left(\mathcal{Z}\right)\right)
\end{eqnarray}
The quantities $j^{+}$ and $j^{-}$ denoted the sets of activating and inhibitory factors for enzyme $j$. 
If a process has no modifying factors, we set $v_{j} = 1$.
There are many possible functional forms for $0\leq f_{ij}\left(\mathcal{Z}\right)\leq 1$. 
However, in this study, each individual transfer function took the form:
\begin{equation}\label{eqn:control-factor}
	f_{i}\left(\mathbf{x}\right) = \frac{\kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}
\end{equation}where $\mathcal{Z}_{j}$ denotes the abundance of the $j$ factor (e.g., metabolite abundance), and $\kappa_{ij}$ and $\eta$ are control parameters. 
The $\kappa_{ij}$ parameter was species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
Applying the general framework to the reduced coagulation network resulted in five ordinary differential equations:
\begin{eqnarray}
	\frac{dx_{1}}{dt} & =& -\left(r_{init}v_{init} + r_{amp}v_{amp}\right)\\
	\frac{dx_{2}}{dt} &=& r_{amp}v_{amp} + r_{init}v_{init} - r_{inh,ATIII}v_{inh,ATIII}\\
	\frac{dx_{3}}{dt} &=& -r_{apc}v_{apc} \\
	\frac{dx_{4}}{dt} &=& r_{apc}v_{apc} \\
	\frac{dx_{5}}{dt} & = & -r_{inh,ATIII}v_{inh,ATIII}
\end{eqnarray}where $\mathbf{x} = \left(fII,FIIa,PC,APC,ATIII\right)^{T}$. 
The terms $r_{*}v_{*}$ in the balance equations denote corrected kinetic expressions for initiation, amplification and inhibition processes. 
The rate of initiation $\bar{r}_{init}$ was modeled as:
\begin{equation}
	\bar{r}_{init} = k_{init}\left(trigger\right)\frac{x_{1}}{K_{init,fII} + x_{1}}
\end{equation}where $k_{init},~K_{init,fII}$ are the rate and saturation constants governing initiation, respectively. 
The rate of initiation was modified by $v_{init}$, the control parameter governing initiation. 
Initiation was sensitive to level of trigger (activator), and TFPI (inhibitor):
\begin{equation}
	v_{init} = \min\left(f^{-}_{init}\left(TFPI\right),f^{+}_{init}\left(trigger\right)\right)
\end{equation}where the transfer functions $f$ took the form of Eqn \eqref{eqn:control-factor}.
The rate of thrombin amplification was given by:
\begin{equation}
	\bar{r}_{amp} = k_{amp}\left(x_{2}\right)\frac{x_{1}}{K_{amp,fII}+x_{1}}
\end{equation}where $k_{amp},~K_{amp,fII}$ denote the rate and saturation constants governing amplification, respectively. 
The amplification control term, which modified amplification rate, was modeled as a combination of multiple inhibition terms and one activation term:
\begin{equation}
	v_{amp} = \min\left(f^{-}_{amp}\left(TFPI\right),f^{-}_{amp}\left(x_{4}\right),f^{+}_{amp}\left(\mathcal{Z}_{amp}\right)\right)
\end{equation}where $\mathcal{Z}_{amp} = fV\times fX\times fVIII\times fIX$. Although, $f^{+}_{amp}\left(\mathcal{Z}_{amp}\right)$ is an activating term, 
we included it in the min integration rule; the factors in $\mathcal{Z}_{amp}$ were essential for amplification (if any of these factors was missing
the amplification reaction would not occur). Thus, the factors in $\mathcal{Z}_{amp}$ were required components, a classification
that we implemented by the min selection rule.
The rate activated protein C formation was given by:
\begin{equation}
	\bar{r}_{apc} = k_{APC,formation}\left(TM\right)\frac{x_{3}}{K_{formation,PC} + x_{3}}
\end{equation}where $k_{APC,formation},~K_{formation,PC}$ denote the rate and saturation constants governing activated protein C formation, respectively and $TM$ denotes
the thrombomodulin abundance. We modeled the control term which governed APC formation as a single thrombin dependent activation term:
\begin{equation}
	v_{apc} = \max\left(f_{apc}^{+}\left(x_{2}\right)\right)
\end{equation}
Lastly, we included direct irreversible inhibition of FIIa by ATIII:
\begin{equation}
	\bar{r}_{inh,ATIII} = k_{ATIII,inhibition}\left(x_{5}x_{2}^{\gamma}\right)
\end{equation}where $\gamma$ was estimated to be $\gamma = 1.26$. For ATIII inhibition of FIIa, the control variables $v_{inh,ATIII}$ was taken to be unity.
The model equations were encoded using the Python programming language and solved using the ODEINT routine of the SciPy module \citep{SCIPY}.
The model files can be downloaded from http://www.varnerlab.org.

\subsection*{Estimation of model parameters from experimental data.}
Model parameters were estimated by minimizing the difference between simulations and experimental thrombin measurements (squared residual):
\begin{equation}\label{eqn:objective-function}
	\min_{\mathbf{k}} \sum_{\tau=1}^{\mathcal{T}}\sum_{j=1}^{\mathcal{S}}\left(\frac{\hat{x}_{j}\left(\tau\right) - x_{j}\left(\tau,\mathbf{k}\right)}{\omega_{j}\left(\tau\right)}\right)^{2}
\end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated 
value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is respect to
time, while the inner summation is with respect to state. We minimized the model residual using Particle swarm optimization (PSO) \citep{PSO}.
PSO uses a \textit{swarming} metaheuristic to explore parameter spaces. 
A strength of PSO is its ability to find the global minimum, even in the presence of potentially many local minima, by communicating the local
error landscape experienced by each particle collectively to the swarm. Thus, PSO acts both as a local and a global search algorithm. 
For each iteration, particles in the swarm compute their local error by evaluating the model equations using their specific parameter vector realization.
From each of these local points, a globally best error is identified. Both the local and global error 
are then used to update the parameter estimates of each particle using the rules:
\begin{eqnarray}
	\mathbf{\Delta}_{i} &=&\theta_{1}\mathbf{\Delta}_{i} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{k}_{i}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G} - \mathbf{k}_{i}\right) \\
	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes local best solution found by particle $i$, and
$\mathcal{G}$ denotes the best solution found over the entire population of particles. The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$. The quality of parameter
estimates was measured using goodness of fit (model residual). The particle swarm optimization routine was implemented in the Python programming language. 
All plots were made using the Matplotlib module of Python \citep{Matplotlib}.
The optimization code can be downloaded from http://www.varnerlab.org.

\subsection*{Global sensitivity analysis of model performance.}
We conducted a global sensitivity analysis to estimate which parameters controlled the performance of the reduced order model. 
We computed the total sensitivity index of each parameter relative to two performance objectives, the peak thrombin time and the area under the thrombin curve 
(thrombin exposure). [FINISH~ME].

@@

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-1-Network.pdf}
\caption{Schematic of the connectivity of the reduced order coagulation model. A trigger compound e.g., TF/FVIIa initiates thrombin production (FIIa) from prothrombin (fII).
Once activated, thrombin catalyzes its own activation (amplification step), as well as its own inhibition via the conversion of protein C to activated protein C (APC).
APC and tissue factor pathway inhibitor (TFPI) inhibit initiation and amplification, while antithrombin III (ATIII) directly inhibits thrombin. 
All inhibition steps, and trigger-induced initiation were modeled using a rule-based approach. 
Likewise, the dependence of amplification on other coagulation factors was also modeled using a rule-based approach. 
The abundance of the highlighted species (in the dashed boxes) was governed by an ordinary differential equation. 
All other species were assumed to be constant.}\label{fig-network}
\end{figure}
